Location and Scale-Invariant Power Transformations for Transforming Data to

### 2025-10-17 manuscript_rmarkdown/manuscript.Rmd
Updated ranking experiment with 10x more sequences.

Three datasets with 10000 sequences each were created to assess transformation
to normality. For each sequence $\mathbf{X}_i$, $n = \lceil 10^\gamma \rceil$
elements were randomly drawn, with $\gamma \sim U\left(1, 4\right)$, 
resulting in sequences with between $10$ and $10000$ elements.

@@

This results in nine types of power transformation. Transformation parameters
were optimised for each sequence. Subsequently the sum of residual errors 
and sum of residual errors of the central portion ($\kappa = 0.80$) were computed
for each sequence after transformation. Then, each method was ranked according 
to the sum of residual errors (data without outliers) or the sum of residual
errors of the central portion (data with outliers). The average rank of each
transformation method was computed over all 10000 sequences in each dataset.
Average ranks for Yeo-Johnson transformations are shown in Table
\ref{tab:comparison_methods_simulation_yeo_johnson}.
Location and shift-invariant Yeo-Johnson transformation ranked best for
all datasets without outliers. The robust variant ranked best in dirty and shifted
datasets with outliers, but not in the clean dataset.
Results for Box-Cox transformations are shown in Table \ref{tab:comparison_methods_simulations_box_cox}.

@@

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 10000 sequences each. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson}
\begin{tabular}{l | l r r r r r r}

@@

none                                  & &         8.51  &         7.60  &         7.33  &         6.15  &         7.44  &         6.60 \\
conventional                          & &         3.70  &         5.70  &         3.87  &         7.06  &         6.49  &         5.93 \\
conventional (z-score norm.)          & &         4.72  &         5.68  &         6.38  &         5.01  &         5.43  &         4.87 \\
conventional (robust scaling)         & &         4.06  &         6.77  &         5.37  &         5.75  &         4.41  &         5.63 \\
Raymaekers-Rousseeuw                  & &         4.50  &         3.20  &         4.01  &         4.13  &         6.27  &         5.65 \\
Raymaekers-Rousseeuw (z-score norm.)  & &         5.27  & \textbf{2.66} &         6.23  &         3.59  &         5.28  &         3.48 \\
Raymaekers-Rousseeuw (robust scaling) & &         4.70  &         2.95  &         5.32  &         3.77  &         4.36  &         3.66 \\
invariant                             & & \textbf{3.13} &         6.77  & \textbf{3.22} &         6.12  & \textbf{2.64} &         6.03 \\
robust invariant                      & &         6.41  &         3.66  &         3.27  & \textbf{3.41} &         2.68  & \textbf{3.10} \\

>>>>

Three datasets with 100000 sequences each were created to assess transformation
to normality. For each sequence $\mathbf{X}_i$, $n = \lceil 10^\gamma \rceil$
elements were randomly drawn, with $\gamma \sim U\left(1, 4\right)$, 
resulting in sequences with between $10$ and $10000$ elements.

@@

This results in nine types of power transformation. Transformation parameters
were optimised for each sequence. Subsequently the sum of residual errors 
and sum of residual errors of the central portion ($\kappa = 0.80$) were computed
for each sequence after transformation. Then, each method was ranked according 
to the sum of residual errors (data without outliers) or the sum of residual
errors of the central portion (data with outliers). The average rank of each
transformation method was computed over all 100000 sequences in each dataset.
Average ranks for Yeo-Johnson transformations are shown in Table
\ref{tab:comparison_methods_simulation_yeo_johnson}.
Location and shift-invariant Yeo-Johnson transformation ranked best for
all datasets without outliers. The robust variant ranked best in dirty and shifted
datasets with outliers, but not in the clean dataset.
Results for Box-Cox transformations are shown in Table \ref{tab:comparison_methods_simulations_box_cox}.

@@

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 sequences each. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson}
\begin{tabular}{l | l r r r r r r}

@@

none                                  & &         8.52  &         7.62  &         7.35  &         6.17  &         7.46  &         6.62 \\
conventional                          & &         3.82  &         5.78  &         3.86  &         7.08  &         6.51  &         5.95 \\
conventional (z-score norm.)          & &         4.70  &         5.75  &         6.44  &         5.07  &         5.48  &         4.93 \\
conventional (robust scaling)         & &         3.83  &         6.61  &         5.36  &         5.72  &         4.41  &         5.59 \\
Raymaekers-Rousseeuw                  & &         4.64  &         3.22  &         4.01  &         4.11  &         6.28  &         5.67 \\
Raymaekers-Rousseeuw (z-score norm.)  & &         5.24  & \textbf{2.66} &         6.25  &         3.62  &         5.29  &         3.50 \\
Raymaekers-Rousseeuw (robust scaling) & &         4.70  &         2.99  &         5.22  &         3.69  &         4.27  &         3.57 \\
invariant                             & & \textbf{3.11} &         6.74  & \textbf{3.24} &         6.13  & \textbf{2.63} &         6.03 \\
robust invariant                      & &         6.44  &         3.64  &         3.27  & \textbf{3.40} &         2.65  & \textbf{3.14} \\
